---
title: "hypothesis_testing"
author: "Yichen Wang"
output:   
  html_document:
    keep_md: yes
---

```{r initialisation, include=FALSE}
knitr::opts_chunk$set(tidy=TRUE,warning = FALSE, message = FALSE)
```

```{r setup}
library(ggplot2)
library(RColorBrewer)
library(nlme)
library(tidyverse)
```

```{r Prepare for the disease effect plotting}

dot_cols = c("#DF5B3F", "#F5B46F", "#70A3C4", "black")
names(dot_cols) = c('Small and large polyp', 'Diminutive polyp', 'ACF', 'Age effect' )


df_sbs_disease = data.frame(matrix(nrow=12,ncol=5))
colnames(df_sbs_disease) = c('rate','type','group','lower_95_CI','upper_95_CI')
```

```{r read file}
database = read.csv('../data/fap_database_withexposure.csv',header=T)
database$type = as.factor(database$type)
database$type = relevel(database$type , ref = "normal")

#add number of APC mutations
database$num_APC = 3-(is.na(database$germlinehit)+is.na(database$somatic1)+is.na(database$somatic2))
database$num_APC = as.numeric(database$num_APC)

database=database[database$coverage>=10,]
database$germlinehit= 'APC'

database_exclude_bystander = database[(database$num_APC>=2 | database$type =='normal' ),]
database_exclude_bystander$group_id[database_exclude_bystander$group_id=='normal'] = database_exclude_bystander$sample[database_exclude_bystander$group_id=='normal']
```

```{r Add control normal colon data, echo=FALSE}
ctrl_database = read.csv('../data/ctrl_database_withexposure.csv',header=T)
ctrl_database$num_APC=0
ctrl_database=ctrl_database[ctrl_database$coverage>=10,]

full_dataset = rbind(database_exclude_bystander,ctrl_database)
full_dataset$germlinehit[is.na(full_dataset$germlinehit)]= 'WT'
full_dataset$germlinehit= as.factor(full_dataset$germlinehit)
full_dataset$germlinehit = relevel(full_dataset$germlinehit , ref = "WT")
pvalues_lst=c()
```


# Whether germline APC mutation leads to incresed sbs burden

```{r SBS total, germline, model selection}
df = full_dataset[full_dataset$type %in%  c('normal'),]
lmm_sbs <- lme(sbstotal_corr~  age:germlinehit , random = ~1|patient/block, data = df, method = "ML")
lmm2_sbs <- lme(sbstotal_corr~ age, random = ~1|patient/block, data = df, method = "ML")
anova(lmm_sbs,lmm2_sbs)
```

```{r SBS total, germline, model}
df = database_exclude_bystander[database_exclude_bystander$type %in%  c('normal'),]
lmm2_sbs <- lme(sbstotal_corr ~ age  , random = ~1|patient/block, data = df, method = "ML")
summary(lmm2_sbs)
intervals(lmm2_sbs, which = "fixed")
pvalues_lst=c(pvalues_lst,summary(lmm2_sbs)$tTable[ , "p-value"])
```


# Whether germline APC mutation leads to incresed Indel burden

```{r ID total, germline, model selection}
df = full_dataset[full_dataset$type %in%  c('normal'),]
lmm_id <- lme(indeltotal_corr ~  age:germlinehit , random = ~1|patient/block, data = df, method = "ML")
lmm2_id <- lme(indeltotal_corr ~ age, random = ~1|patient/block, data = df, method = "ML")
anova(lmm_id,lmm2_id)

```


```{r ID total, germline, model}
df = database_exclude_bystander[database_exclude_bystander$type %in%  c('normal'),]
lmm2_id <- lme(indeltotal_corr ~ age  , random = ~1|patient, data = df, method = "ML")
summary(lmm2_id)

fixed.m1 <- data.frame(fixef(lmm2_id))
intervals(lmm2_id, which = "fixed")
pvalues_lst=c(pvalues_lst,summary(lmm2_id)$tTable[ , "p-value"])
```

```{r Calcultae P-adjusted for SBS and ID burden in normal crypts}
p_corrected =  p.adjust(pvalues_lst, method = "BH")
p_table = data.frame(p=pvalues_lst,p_adj=p_corrected)
```

## Whetehr bystander crypts are the same as normal
```{r construct dataset for bystander crypts, echo=FALSE}
database_bystander = database[(database$num_APC<2 & database$type %in% c('normal','acf')),]
database_bystander$group_id[database_bystander$group_id=='normal'] = database_bystander$sample[database_bystander$group_id=='normal']
ctrl_database = read.csv('../data/ctrl_database_withexposure.csv',header=T)
ctrl_database$num_APC=0
ctrl_database=ctrl_database[ctrl_database$coverage>=10,]

full_dataset_bystander = rbind(database_bystander,ctrl_database)
full_dataset_bystander$germlinehit[is.na(full_dataset_bystander$germlinehit)]= 'WT'
full_dataset_bystander$germlinehit= as.factor(full_dataset_bystander$germlinehit)
full_dataset_bystander$germlinehit = relevel(full_dataset_bystander$germlinehit , ref = "WT")

```


```{r SBS total, bystander}
df = full_dataset_bystander
lmm_sbs <- lme(sbstotal_corr~ age+germlinehit, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_sbs)
```

```{r ID total, bystander}
df = full_dataset_bystander
lmm_id <- lme(indeltotal_corr~ age+germlinehit, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_id)
```


# Whether ACF leads to incresed sbs burden
```{r SBS18, ACF}
pvalues_lst=c()
df = full_dataset[full_dataset$type %in%  c('normal','acf'),]
lmm_sbs18 <- lme(SBS18~ age+type, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_sbs18)
pvalues_lst=c(pvalues_lst,summary(lmm_sbs18)$tTable[ , "p-value"])
intervals(lmm_sbs18, which = "fixed")
```


```{r SBS1, ACF}
df = full_dataset[full_dataset$type %in%  c('normal','acf'),]
lmm_sbs1 <- lme(SBS1~ age+type, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_sbs1)
pvalues_lst=c(pvalues_lst,summary(lmm_sbs1)$tTable[ , "p-value"])
intervals(lmm_sbs1, which = "fixed")
```

```{r SBS5, ACF}
df = full_dataset[full_dataset$type %in%  c('normal','acf'),]
lmm_sbs5 <- lme(SBS5~ age+type, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_sbs5)
pvalues_lst=c(pvalues_lst,summary(lmm_sbs5)$tTable[ , "p-value"])
intervals(lmm_sbs5, which = "fixed")
```

```{r SBS total, ACF}
df = full_dataset[full_dataset$type %in%  c('normal','acf'),]
lmm_sbs <- lme(sbstotal_corr~ age+type, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_sbs)
pvalues_lst=c(pvalues_lst,summary(lmm_sbs)$tTable[ , "p-value"])
intervals(lmm_sbs, which = "fixed")
```

# Whether ACF leads to incresed Indel burden
```{r ID1, ACF}
df = full_dataset[full_dataset$type %in%  c('normal','acf'),]
lmm_id1 <- lme(ID1~  age +type  , random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_id1)
pvalues_lst=c(pvalues_lst,summary(lmm_id1)$tTable[ , "p-value"])
intervals(lmm_id1, which = "fixed")
```

```{r ID2, ACF}
df = full_dataset[full_dataset$type %in%  c('normal','acf'),]
lmm_id2 <- lme(ID2 ~  age +type , random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_id2)
pvalues_lst=c(pvalues_lst,summary(lmm_id2)$tTable[ , "p-value"])
intervals(lmm_id2, which = "fixed")
```

```{r ID5, ACF}
df = full_dataset[full_dataset$type %in%  c('normal','acf'),]
lmm_id5 <- lme(ID5 ~ age+type, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_id5)
pvalues_lst=c(pvalues_lst,summary(lmm_id5)$tTable[ , "p-value"])
intervals(lmm_id5, which = "fixed")
```

```{r ID total, ACF}
df = full_dataset[full_dataset$type %in%  c('normal','acf'),]
lmm_id <- lme(indeltotal_corr~  age+type , random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_id)
pvalues_lst=c(pvalues_lst,summary(lmm_id)$tTable[ , "p-value"])
intervals(lmm_id, which = "fixed")
```
```{r adjusted P table, ACF}
p_corrected_acf =  p.adjust(pvalues_lst, method = "BH")
p_table_acf = data.frame(p=pvalues_lst,p_adj=p_corrected_acf)
```

# Whether Polyp leads to incresed sbs burden
```{r SBS18, diminutive polyps}
pvalues_lst=c()
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf'),]
lmm_sbs18 <- lme(SBS18~ type+age, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_sbs18)
pvalues_lst=c(pvalues_lst,summary(lmm_sbs18)$tTable[ , "p-value"])
intervals(lmm_sbs18, which = "fixed")
```


```{r SBS1, diminutive polyps}
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf'),]
lmm_sbs1 <- lme(SBS1~ age+type, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_sbs1)
pvalues_lst=c(pvalues_lst,summary(lmm_sbs1)$tTable[ , "p-value"])
intervals(lmm_sbs1, which = "fixed")
```


```{r SBS5, diminutive polyps}
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf'),]
lmm_sbs5 <- lme(SBS5~ age+type, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_sbs5)
pvalues_lst=c(pvalues_lst,summary(lmm_sbs5)$tTable[ , "p-value"])
intervals(lmm_sbs5, which = "fixed")
```


```{r SBS total, diminutive polyps}
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf'),]
lmm_sbs <- lme(sbstotal_corr~ age+type, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_sbs)
pvalues_lst=c(pvalues_lst,summary(lmm_sbs)$tTable[ , "p-value"])
intervals(lmm_sbs, which = "fixed")
```


# Whether polyps leads to incresed Indel burden
```{r ID1, diminutive polyps}
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf'),]
lmm_id1 <- lme(ID1~  age +type  , random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_id1)
pvalues_lst=c(pvalues_lst,summary(lmm_id1)$tTable[ , "p-value"])
intervals(lmm_id1, which = "fixed")
```


```{r ID2, diminutive polyps}
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf'),]
lmm_id2 <- lme(ID2 ~  age +type , random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_id2)
pvalues_lst=c(pvalues_lst,summary(lmm_id2)$tTable[ , "p-value"])
intervals(lmm_id2, which = "fixed")
```


```{r ID5, diminutive polyps}
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf'),]
lmm_id5 <- lme(ID5 ~ age+type, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_id5)
pvalues_lst=c(pvalues_lst,summary(lmm_id5)$tTable[ , "p-value"])
intervals(lmm_id5, which = "fixed")
```


```{r ID total, diminutive polyps}
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf'),]
lmm_id <- lme(indeltotal_corr~  age+type , random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_id)
pvalues_lst=c(pvalues_lst,summary(lmm_id)$tTable[ , "p-value"])
intervals(lmm_id, which = "fixed")
```


```{r adjuested P table, diminutive polyps}
p_corrected_polyp =  p.adjust(pvalues_lst, method = "BH")
p_table_polyp = data.frame(p=pvalues_lst,p_adj=p_corrected_polyp)
```

# Whether large polyp leads to incresed sbs burden
```{r SBS18, large polyps}
pvalues_lst=c()
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf','large polyp'),]
lmm_sbs18 <- lme(SBS18~ type+age, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_sbs18)
pvalues_lst=c(pvalues_lst,summary(lmm_sbs18)$tTable[ , "p-value"])
intervals(lmm_sbs18, which = "fixed")
```

```{r SBS1, large polyps}
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf','large polyp'),]
lmm_sbs1 <- lme(SBS1~ age+type, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_sbs1)
pvalues_lst=c(pvalues_lst,summary(lmm_sbs1)$tTable[ , "p-value"])
intervals(lmm_sbs1, which = "fixed")
```

```{r SBS5, large polyps}
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf','large polyp'),]
lmm_sbs5 <- lme(SBS5~ age+type, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_sbs5)
pvalues_lst=c(pvalues_lst,summary(lmm_sbs5)$tTable[ , "p-value"])
intervals(lmm_sbs5, which = "fixed")
```


```{r SBS total, large polyps}
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf','large polyp'),]
lmm_sbs <- lme(sbstotal_corr~ age+type, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_sbs)
pvalues_lst=c(pvalues_lst,summary(lmm_sbs)$tTable[ , "p-value"])
intervals(lmm_sbs, which = "fixed")
```

# Whether large polyp leads to incresed Indel burden
```{r ID1, large polyps}
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf','large polyp'),]
lmm_id1 <- lme(ID1~  age +type  , random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_id1)
pvalues_lst=c(pvalues_lst,summary(lmm_id1)$tTable[ , "p-value"])
intervals(lmm_id1, which = "fixed")
```

```{r ID2, large polyps}
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf','large polyp'),]
lmm_id2 <- lme(ID2 ~  age +type , random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_id2)
pvalues_lst=c(pvalues_lst,summary(lmm_id2)$tTable[ , "p-value"])
intervals(lmm_id2, which = "fixed")
```

```{r ID5, large polyps}
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf','large polyp'),]
lmm_id5 <- lme(ID5 ~ age+type, random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_id5)
pvalues_lst=c(pvalues_lst,summary(lmm_id5)$tTable[ , "p-value"])
intervals(lmm_id5, which = "fixed")
```


```{r ID total, large polyps}
df = full_dataset[full_dataset$type %in%  c('normal','polyp','acf','large polyp'),]
lmm_id <- lme(indeltotal_corr~  age+type , random = ~1|patient/block/group_id, data = df, method = "ML")
summary(lmm_id)
pvalues_lst=c(pvalues_lst,summary(lmm_id)$tTable[ , "p-value"])
intervals(lmm_id, which = "fixed")
```

```{r adjusted P table, large polyps}
p_corrected_large_polyp =  p.adjust(pvalues_lst, method = "BH")
p_table_large_polyp = data.frame(p=pvalues_lst,p_adj=p_corrected_large_polyp)
```



